Finiteness in Cubical Type Theory

### 2020-06-29 sections/finiteness-predicates.tex
kuratowski

So our tour through the finiteness predicates takes us next to one which allows
the circle to be called finite: manifest enumerability.
\begin{definition}[Manifest Enumerability]
  Manifest enumerability is an enumeration predicate like Bishop finiteness or
  split enumerability with the only difference being a propositionally truncated
  membership proof.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/ManifestEnumerable/Container.tex]{manifest-enum-def}
  \end{agdalisting}
\end{definition}
As a function-based definition, this predicate represents surjections.

\begin{definition}[Surjections] \label{surjections}
  We define proper surjections (not split surjections) here \cite[definition
  4.6.1]{hottbook}.
  \begin{alignat}{3}
    &\text{surj}(f)             &&\coloneqq \Pi(y : B) , \lVert \text{fib}_f(y) \rVert \label{surj-eqn} \\
    &A \twoheadrightarrow B     &&\coloneqq \Sigma (f : A \rightarrow B) , \text{surj}(f) \label{surj-arrow-eqn}
  \end{alignat}
\end{definition}
\begin{lemma}
  Manifest enumerability is equivalent to a surjection from a finite prefix of
  the natural numbers.
  \begin{equation}
    \mathcal{E}(A) \simeq \Sigma(n : \mathbb{N}) , (\AgdaDatatype{Fin}\;n \twoheadrightarrow A)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
    && \text{def.~\ref{split-enum-def} }(\mathcal{E})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \lVert \text{fib}_{\text{snd}(\mathit{xs})}(x) \rVert
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{surj-eqn} (surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \AgdaDatatype{Fin} \rrbracket (A)) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\AgdaDatatype{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \AgdaDatatype{Fin}\;n) , A) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \AgdaDatatype{Fin}\;n \rightarrow A) , \text{surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \AgdaDatatype{Fin}\;n \twoheadrightarrow A )
    && \text{eqn.~\ref{surj-arrow-eqn} } (\twoheadrightarrow) \; \qedhere
  \end{align*}
\end{proof}
\subsection{Instances for Non-Set Types}
The truncation has another very important implication: it means that the
predicate doesn't provide decidable equality on the underlying type.
Remember, this is how we knew that our previous predicates wouldn't allow for
non-set types: because they implied decidable equality, they also implied that
all conforming types had homotopy levels of at most 2. \todo{Are we doing the
  homotopy levels starting from -2 thing or from 0?}
This suggests that non-set types like the circle could conform to this
finiteness predicate.
\begin{lemma}
  The circle \(S^1\) is manifestly enumerable.
\end{lemma}
\begin{proof}
  The support list firstly is a list containing the point constructor for the
  circle.
  Since the cover proof is truncated, we need only consider the point
  constructors of the circle: as such, the cover proof is essentially the same
  as the one for \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}(\top)\).
\end{proof}
\subsection{Relation To Split Enumerability}
It is trivially easy to construct a proof that any split enumerable type is
manifest enumerable: we simply truncate the membership proof.
Going the other was requires us to extract a non-truncated proof from a
truncated one.
This proof relies on the following lemma:
\begin{lemma}
  We can ``recompute'' a truncated proof given a decision over a proof of the
  same type.
  \begin{equation}
    \frac{\lVert A \rVert \; \; \; \mathbf{Dec}(A)}{A}
  \end{equation}
\end{lemma}
\begin{proof}
  We proceed by case-analysis over the decision over \(A\).
  In the case where \(A\) is proven, we are done.
  In the case where \(A\) is disproven, we use lemma~\ref{prop-refute} to
  derive impossibility.
\end{proof}

@@

  We do this using the previous recompute lemma: that tells us that all we need
  to construct is a decision for \(x \in \mathit{xs}\), and it will be able to
  derive the proof itself.
  Such a decision procedure is not difficult to construct: for any value \(x\)
  and list \(\mathit{xs}\), we proceed through the list \(\mathit{xs}\), testing
  if \(x\) is equal to any of its contents.
  If it is, we return that we have proven the goal, and that \(x\) is indeed
  present in \(\mathit{xs}\).
  Otherwise, we know that \(x\) cannot be in \(\mathit{xs}\) (since we've tested
  every value), so we return that the goal has been disproven.
\end{proof}
\section{Kuratowski Finiteness} \label{kuratowski}
The one big missing definition of finiteness to cover is \emph{Kuratowski}
finiteness.
While it's quite important, it's also quite different from the definitions we've
seen so far.
It starts with an encoding of the free join semilattice.
\begin{definition}[Free Join Semilattice]
  \(\mathcal{K}(A)\) is the free join semilattice, or, alternatively, the type
  of Kuratowski-finite subsets of \(A\).
  \begin{inductivetype}{\mathcal{K}(A)}
    \inductivetypeclause{\left[ \right]}{\mathcal{K}(A)}
    \inductivetypeclause{\wc \dblcolon \wc}{A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A)}
    \inductivetypeclause{\text{com}}{\Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{dup}}{\Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{trunc}}{\Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q}
  \end{inductivetype}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

@@

Next, we need a way to say that an entire type is Kuratowski finite.
For that, we will need to define membership of \(\mathcal{K}\).
\begin{definition}[Membership of \(\mathcal{K}\)]
  Membership is defined by pattern-matching on \(\mathcal{K}\).
  The two point constructors are handled like so:
  \begin{equation}
    \begin{alignedat}{2}
      x \in&& \; []                      &\coloneqq \bot ; \\
      x \in&& \; y \dblcolon \mathit{ys} &\coloneqq \lVert x \equiv y \uplus x \in \mathit{ys} \rVert ;
    \end{alignedat}
  \end{equation}
  The \(\text{com}\) and \(\text{dup}\) constructors are handled by proving that
  the truncated form of \(\uplus\) is itself commutative and idempotent.
  The type of propositions is itself a set, satisfying the \(\text{trunc}\)
  constructor.
\end{definition}
Finally, we have enough background to define Kuratowski finiteness.
\begin{definition}[Kuratowski Finiteness]
  \begin{equation}
    \mathcal{K}^{f}(A) = \Sigma {(\mathit{xs} : \mathcal{K}(A))} , \Pi (x : A) , x \in \mathit{xs}
  \end{equation}
\end{definition}

>>>>

So our tour through the finiteness predicates takes us next to one which allows
the circle to be called finite: manifest enumerability.
\begin{definition}[Manifest Enumerability]
  Manifest enumerability is an enumeration predicate like Bishop finiteness or
  split enumerability with the only difference being a propositionally truncated
  membership proof.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/ManifestEnumerable/Container.tex]{manifest-enum-def}
  \end{agdalisting}
\end{definition}

It might not be immediately clear why this definition of enumerability allows
the circle to conform while the others do not.
The crux of the issue was that the cover proofs of the previous definitions
didn't just tell us that some element was in the support list, they told us
\emph{where} it was in the support list.
From the position we were able to derive decidable equality: that position is
precisely what's hidden in manifest enumerability.

And indeed this means that the circle is manifestly enumerable:
\begin{lemma}
  The circle \(S^1\) is manifestly enumerable.
\end{lemma}
\begin{proof}
  The support list firstly is a list containing the point constructor for the
  circle.
  Since the cover proof is truncated, we need only consider the point
  constructors of the circle: as such, the cover proof is essentially the same
  as the one for \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}(\top)\).
\end{proof}
\subsection{Surjections}
We already say that split enumerability was the listed form of a split
surjection: manifest enumerability is the listed form of a proper surjection.
\begin{definition}[Surjections] \label{surjections}
  We define proper surjections (not split surjections) here \cite[definition
  4.6.1]{hottbook}.
  \begin{alignat}{3}
    &\text{surj}(f)             &&\coloneqq \Pi(y : B) , \lVert \text{fib}_f(y) \rVert \label{surj-eqn} \\
    &A \twoheadrightarrow B     &&\coloneqq \Sigma (f : A \rightarrow B) , \text{surj}(f) \label{surj-arrow-eqn}
  \end{alignat}
\end{definition}
\begin{lemma}
  Manifest enumerability is equivalent to a surjection from a finite prefix of
  the natural numbers.
  \begin{equation}
    \mathcal{E}(A) \simeq \Sigma(n : \mathbb{N}) , (\AgdaDatatype{Fin}\;n \twoheadrightarrow A)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
    && \text{def.~\ref{split-enum-def} }(\mathcal{E})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \lVert \text{fib}_{\text{snd}(\mathit{xs})}(x) \rVert
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{surj-eqn} (surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \AgdaDatatype{Fin} \rrbracket (A)) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\AgdaDatatype{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \AgdaDatatype{Fin}\;n) , A) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \AgdaDatatype{Fin}\;n \rightarrow A) , \text{surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \AgdaDatatype{Fin}\;n \twoheadrightarrow A )
    && \text{eqn.~\ref{surj-arrow-eqn} } (\twoheadrightarrow) \; \qedhere
  \end{align*}
\end{proof}
\subsection{Relation To Split Enumerability}
It is trivially easy to construct a proof that any split enumerable type is
manifest enumerable: we simply truncate the membership proof.
Going the other was requires us to extract a non-truncated proof from a
truncated one.
This proof relies on the following lemma:
\begin{lemma}
  We can ``recompute'' a truncated proof given a decision over a proof of the
  same type.
  \begin{equation}
    \frac{\lVert A \rVert \; \; \; \mathbf{Dec}(A)}{A}
  \end{equation}
\end{lemma}
\begin{proof}
  We proceed by case-analysis over the decision over \(A\).
  In the case where \(A\) is proven, we are done.
  In the case where \(A\) is disproven, we use lemma~\ref{prop-refute} to
  derive impossibility.
\end{proof}

@@

  We do this using the previous recompute lemma: that tells us that all we need
  to construct is a decision for \(x \in \mathit{xs}\), and it will be able to
  derive the proof itself.
  Such a decision procedure is not difficult to construct: for any value \(x\)
  and list \(\mathit{xs}\), we proceed through the list \(\mathit{xs}\), testing
  if \(x\) is equal to any of its contents.
  If it is, we return that we have proven the goal, and that \(x\) is indeed
  present in \(\mathit{xs}\).
  Otherwise, we know that \(x\) cannot be in \(\mathit{xs}\) (since we've tested
  every value), so we return that the goal has been disproven.
\end{proof}
\section{Kuratowski Finiteness} \label{kuratowski} \todo{better intro}
The one big missing definition of finiteness to cover is \emph{Kuratowski}
finiteness.
While it's quite important, it's also quite different from the definitions we've
seen so far.
It starts with an encoding of the free join semilattice.
\begin{definition}[Free Join Semilattice]
  \(\mathcal{K}(A)\) is the free join semilattice, or, alternatively, the type
  of Kuratowski-finite subsets of \(A\).
  \begin{agdalisting}
    \ExecuteMetaData[agda/Algebra/Construct/Free/Semilattice/Definition.tex]{kuratowski-def}
  \end{agdalisting}

  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

@@

Next, we need a way to say that an entire type is Kuratowski finite.
For that, we will need to define membership of \(\mathcal{K}\).
\begin{definition}[Membership of \(\mathcal{K}\)]
  Membership is defined by pattern-matching on \(\mathcal{K}\).
  The two point constructors are handled like so: \todo{to agda}
  \begin{equation}
    \begin{alignedat}{2}
      x \in&& \; []                      &\coloneqq \bot ; \\
      x \in&& \; y \dblcolon \mathit{ys} &\coloneqq \lVert x \equiv y \uplus x \in \mathit{ys} \rVert ;
    \end{alignedat}
  \end{equation}
  The \(\text{com}\) and \(\text{dup}\) constructors are handled by proving that
  the truncated form of \(\uplus\) is itself commutative and idempotent.
  The type of propositions is itself a set, satisfying the \(\text{trunc}\)
  constructor.
\end{definition}
Finally, we have enough background to define Kuratowski finiteness.
\begin{definition}[Kuratowski Finiteness] \phantom{break}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/Kuratowski.tex]{kuratowski-finite-def}
  \end{agdalisting}
\end{definition}
